Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2019-12-11 article-information-2019.tex
edits

%=================================================================
% Full title of the paper (Capitalized)
\Title{Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Disparate Impact Testing}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on realistic, publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using straightforward group fairness measures. By combining innovations in interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Even if responsible AI feels like a somewhat broad and amorphous notion, at least one aspect should be crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate, noncompliant, and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. Though this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), this text does propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems for high-stakes, human-centered, or regulated ML applications.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the U.S., interpretable models, explanations, disparate impact testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (E.U.) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this manuscript with a brief outline of proposed additional steps to increase human trust and understanding in ML.

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository for this text; utilized \& useful Python packages -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, this manuscript compares the performance of more interpretable constrained ML models and less interpretable unconstrained ML models on simulated data and on collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a more realistic template workflow for future users of similar methods and tools.

Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods include measures of DI with marginal effects, the adverse impact ratio (AIR), and the standardized mean difference (SMD) \cite{feldman2015certifying}, \cite{cohen2013statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment\cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated with this text.

@@

	\begin{itemize}[leftmargin=*,labelsep=5.8mm]
		\item A type of ML model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for discrimination testing is denoted as $g$.
	\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Monotonically Constrained Gradient Boosting Machine}\label{ssec:mgbm}

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints. The MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, but each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

@@

\noindent In addition to $\mathcal{L}_{b}$, $g^\text{MGBM}$ training is characterized by additional splitting rules and constraints on tree node weights. Each binary splitting rule, $\theta_{b,j,k} \in \Theta_b$, is associated with a feature, $X_j$, is the $k$-th split associated with $X_j$ in $T_b$, and results in left and right child nodes with a numeric weights, $\{w_{b,j,k,L}, w_{b,j,kR}\}$. For terminal nodes, $\{w_{b,j,k,L}, w_{b,j,kR}\}$ can be direct numeric components of some $g^\text{MGBM}$ prediction. For two values of some feature $X_j$, $x^{\alpha}_j \le x^{\beta}_j$, where the prediction for each value results in $T_b(x^{\alpha}_j; \Theta_b) = w_\alpha$ and $T_b(x^{\beta}_j; \Theta_b) = w_\beta$, $\Theta_b$ is restricted to be positive monotonic w.r.t. $X_j$ by the following rules and constraints.

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item For the first and highest split in $T_b$ involving $X_j$, any $\theta_{b,j,0}$ resulting in the left child weight being greater than the right child weight, $T(x_j; \theta_{b,j,0}) = \{w_{b,j,0,L}, w_{j,0,R}\}$ where $w_{b,j,0,L} > w_{b,j,0,R}$, is not considered. 
\item For any subsequent left child node involving $X_j$, any $\theta_{b,j, k\ge1}$ resulting in $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ where $w_{b,j,k\ge1,L} > w_{b,j,k\ge1,R}$, is not considered.
\item Moreover, for any subsequent left child node involving $X_j$, $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$, $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ are bound by the parent set of node weights, $\{w_{b,j,k-1,L}, w_{b,j,k-1, R}\}$, such that $ \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \le \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\item (1) and (2) are also applied to all right child nodes, except that for right child nodes $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \ge \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\end{enumerate}

\noindent Note that for any one $X_j$ and $T_b \in g^{\text{MGBM}}$ left subtrees will alway produce lower predictions than right subtrees, and that any $g^{\text{MGBM}}(\mathbf{x})$ is an addition of each $T_b$ output, with the application of a monotonic logit or softmax link function for classification problems. Moreover, each tree's root node corresponds to some constant node weight that by definition obeys monotonicity constraints, $ T(x^{\alpha}_j; \theta_{b,0}) = T(x^{\beta}_j; \theta_{b,j,0}) = w_{b,0}$. Together these additional splitting rules and node weight constraints ensure that $g^{\text{MGBM}}(x^{\alpha}_j)  \le g^{\text{MGBM}}(x^{\beta}_j) ~\forall ~x^{\alpha}_j \le x^{\beta}_j \in X_j$. For a negative monotonic constraint, i.e. $g^{\text{MGBM}}(x^{\alpha}_j)  \ge g^{\text{MGBM}}(x^{\beta}_j) ~\forall ~x^{\alpha}_j \le x^{\beta}_j \in X_j$, left and right splitting rules and node weight constraints are switched.\\

Herein, two $g^\text{MGBM}$ models are trained.  One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using the sign of the Pearson correlation between each $X_j$ and $y$, random grid search is used to determine other hyperparameters, and five-fold cross validation and test partitions are used for model assessment. For exact parameterization of the MGBM models, see the software resources referenced in Subsection \ref{ssec:soft}. Also consider that MGBM models with one-dimensional monotonicity constraints between some $X_j$ and $\hat{Y}$ likely restrict non-monotonic interactions between multiple $X_j$. Moreover, if monotonicity constraints are not applied to all $X_j \in \mathbf{X}$, any strong non-monotonic signal in training data associated with some important $X_j$ maybe forced onto some other arbitrary unconstrained $X_j$ under some $g^{MGBM}$ models, compromising the end goal of interpretability.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bare a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e. GAMs which consider main effects and a small number of 2-way interactions and incorporate boosting in their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum^J_{j=0}\beta_{k, j}x_j$.

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=1}^K\gamma_k n_k(\sum^J_{j=1}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

@@

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item The first and deepest meta-layer, composed of $K$ linear $\sum_j\beta_{k,j}x_j$ hidden units, is known as the \textit{projection layer} and is fully connected to each input feature, $X_j$.
\item The second meta-layer contains $K$ hidden and separate $n_k$ ridge functions, or \textit{subnetworks}. Each $n_k$ is a neural network, which can be parameterized to suite a given modeling task. To facilitate direct visualization, the input to each subnetwork is the 1-dimensional output of its associated projection layer hidden unit, $\sum_j\beta_{k,j}x_j$.
\item The output meta-layer, called the \textit{combination layer}, is another linear unit comprised of a global bias term, $\mu_0$, and the $K$ weighted 1-dimensional outputs of each subnetwork, $\gamma_kn_k(\sum_j\beta_{k,j}x_j)$. Again, subnetwork output is restricted to 1-dimension for visualization purposes.
\end{enumerate}

Here, each $g^\text{XNN}$ is trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition.  $L_1$ regularization is applied to both the projection and combination layers to induce a sparse and interpretable model, where each $n_k$ subnetwork and corresponding combination layer $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models are highly sensitive to weight initialization and manual feature selection is informed by Shapley feature importance from the $g^\text{MGBM}$. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft}. Be aware that $g^{XNN}$ model architectures may restrict important interactions, appear highly sensitive to initialization routines, and require manual and judicious feature selection due to burdensome training times.

@@

PD plots are a widely-used method for describing and plotting the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g. inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions of a complex model for some $X_j$ \cite{ice_plots}.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using adverse impact ratio (AIR), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. While this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems for high-stakes, human-centered, or regulated ML applications.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an appealable decision-making framework, tests for DI and other social discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this manuscript with a brief summary of the outlined methods, materials, results, and discussion.

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: AIR, SMD, and other group fairness metrics -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository associated with Sections \ref{sec:m_and_m} and \ref{sec:res} -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis.

Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a more realistic template workflow for future users of similar methods and tools. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods include measures of DI with marginal effects, AIR and SMD \cite{feldman2015certifying}, \cite{cohen2013statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated delineated in Subsection \ref{ssec:soft}.

@@

	\begin{itemize}[leftmargin=*,labelsep=5.8mm]
		\item A type of ML model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for discrimination is denoted as $g$.
	\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Monotonic Gradient Boosting Machine}\label{ssec:mgbm}

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each $X_j$ and $\mathbf{y}$ independently. The MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

@@

\noindent In addition to $\mathcal{L}_{b}$, $g^\text{MGBM}$ training is characterized by additional splitting rules and constraints on tree node weights. Each binary splitting rule, $\theta_{b,j,k} \in \Theta_b$, is associated with a feature, $X_j$, is the $k$-th split associated with $X_j$ in $T_b$, and results in left and right child nodes with a numeric weights, $\{w_{b,j,k,L}, w_{b,j,k,R}\}$. For terminal nodes, $\{w_{b,j,k,L}, w_{b,j,k,R}\}$ can be direct numeric components of some $g^\text{MGBM}$ prediction. For two values of some feature $X_j$, $x^{\alpha}_j \le x^{\beta}_j$, $g^{\text{MGBM}}$ is positive monotonic with respect to some $X_j$ if $g^{\text{MGBM}}(x^{\alpha}_j) \le g^{\text{MGBM}}(x^{\beta}_j) ~\forall ~x^{\alpha}_j \le x^{\beta}_j \in X_j$. The following rules and constraints ensure positive monotonicity in $\Theta_b$, where the prediction for each value results in $T_b(x^{\alpha}_j; \Theta_b) = w_\alpha$ and $T_b(x^{\beta}_j; \Theta_b) = w_\beta$.

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item For the first and highest split in $T_b$ involving $X_j$, any $\theta_{b,j,0}$ resulting in the left child weight being greater than the right child weight, $T(x_j; \theta_{b,j,0}) = \{w_{b,j,0,L}, w_{b,j,0,R}\}$ where $w_{b,j,0,L} > w_{b,j,0,R}$, is not considered. 
\item For any subsequent left child node involving $X_j$, any $\theta_{b,j, k\ge1}$ resulting in $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ where $w_{b,j,k\ge1,L} > w_{b,j,k\ge1,R}$, is not considered.
\item Moreover, for any subsequent left child node involving $X_j$, $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$, $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ are bound by the associated $\theta_{b,j,k-1}$ set of node weights, $\{w_{b,j,k-1,L}, w_{b,j,k-1, R}\}$, such that $ \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} < \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\item (1) and (2) are also applied to all right child nodes, except that for right child nodes $ w_{b,j,k,L} \le w_{b,j,k,R}$ and $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \ge \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\end{enumerate}

\noindent Note that for any one $X_j$ and $T_b \in g^{\text{MGBM}}$ left subtrees will alway produce lower predictions than right subtrees, and that any $g^{\text{MGBM}}(\mathbf{x})$ is an addition of each $T_b$ output, with the application of a monotonic logit or softmax link function for classification problems. Moreover, each tree's root node corresponds to some constant node weight that by definition obeys monotonicity constraints, $ T(x^{\alpha}_j; \theta_{b,0}) = T(x^{\beta}_j; \theta_{b,0}) = w_{b,0}$. Together these additional splitting rules and node weight constraints ensure that $g^{\text{MGBM}}(x^{\alpha}_j)  \le g^{\text{MGBM}}(x^{\beta}_j) ~\forall ~x^{\alpha}_j \le x^{\beta}_j \in X_j$. For a negative monotonic constraint, i.e. $g^{\text{MGBM}}(x^{\alpha}_j)  \ge g^{\text{MGBM}}(x^{\beta}_j) ~\forall ~x^{\alpha}_j \le x^{\beta}_j \in X_j$, left and right splitting rules and node weight constraints are switched. Also consider that MGBM models with independent monotonicity constraints between some $X_j$ and $\mathbf{y}$ likely restrict non-monotonic interactions between multiple $X_j$. Moreover, if monotonicity constraints are not applied to all $X_j \in \mathbf{X}$, any strong non-monotonic signal in training data associated with some important $X_j$ maybe forced onto some other arbitrary unconstrained $X_j$ under some $g^{MGBM}$ models, compromising the end goal of interpretability.

Herein, two $g^\text{MGBM}$ models are trained.  One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and five-fold cross validation and test partitions are used for model assessment. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bare a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e. GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum_j\beta_{k,j}x_j$.

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J=\mathcal{P}-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

@@

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item The first and deepest meta-layer, composed of $K$ linear $\sum_j\beta_{k,j}x_j$ hidden units, is known as the \textit{projection layer} and is fully connected to each input feature, $X_j$. Each hidden unit in the projection layer may optionally include a bias term.
\item The second meta-layer contains $K$ hidden and separate $n_k$ ridge functions, or \textit{subnetworks}. Each $n_k$ is a neural network, which can be parameterized to suit a given modeling task. To facilitate direct interpretation and visualization, the input to each subnetwork is the 1-dimensional output of its associated projection layer hidden unit, $\sum_j\beta_{k,j}x_j$. Each $n_k$ can contain several bias terms.
\item The output meta-layer, called the \textit{combination layer}, is another linear unit comprised of a global bias term, $\mu_0$, and the $K$ weighted 1-dimensional outputs of each subnetwork, $\gamma_kn_k(\sum_j\beta_{k,j}x_j)$. Again, subnetwork output is restricted to 1-dimension for interpretation and visualization purposes.
\end{enumerate}

Here, each $g^\text{XNN}$ is trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition. $L_1$ regularization is applied to both the projection and combination layers to induce a sparse and interpretable model, where each $n_k$ subnetwork and corresponding combination layer $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft}.

@@

PD plots are a widely-used method for describing and plotting the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g. inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}.
